### 2020-04-14 serial.tex
a bit more

Since the emergence of the novel coronavirus disease (COVID-19), a significant amount of research has focused on estimating its reproduction number $\mathcal R$.
The reproduction number, which is defined as the average number of secondary cases caused by a primary case, allows us to predict the extent to which a disease will spread in the population and the amount of intervention to prevent an outbreak.
Since the reproduction number cannot be measured directly, it is often estimated from the observed exponential growth rate using generation- and serial-interval distributions.

The generation interval is defined as the time between when an individual (infector) is infected and when an individual infects another person (infectee).
On the other hand, the serial interval is defined as the time between when an infector and an infectee become symptomatic.
% Despite having similar definitions, generation and serial intervals are different because serial intervals rely on the generation intervals as well as the incubation periods of an infector and an infectee.
Previous studies have often alluded to differences in their variances, despite having seemingly identical expected values, and noted that using serial-interval distribution can give different estimates of $\mathcal R$.
Even though these distinctions were first made over a decade ago \citep{svensson2007note}, 
the need for a better conceptual and theoretical framework for understanding their differences is becoming clearer as the COVID-19 pandemic unfolds:
Researchers continue to rely on both generation and serial intervals to estimate $\mathcal R$ for COVID-19 without making a clear distinction, and some even mislabel them.

@@

During the initial phase of an epidemic, the proprotion susceptible remains constant ($S(t) = S(0)$) and incidence of infection grows exponentially: $i(t)=i_0\exp(rt)$.
Then, we can estimate the reproduction number from the exponential growth rate $r$ via the Euler-Lotka equation:
\begin{equation}
\frac{1}{\mathcal R} = \int_0^\infty \exp(-r\tau) g(\tau) \mathrm{d} \tau.
\end{equation}
Like forward generation-interval distributions, 
forward serial-interval distributions describe the renewal process of symptomatic cases.
Therefore, the forward serial-interval distribution $f_{\tiny\textrm{exp}}(\tau)$ during the exponential growth phase provides the identical $r$--$\mathcal R$ link as the intrinsic generation-interval distribution:
\begin{equation}
\frac{1}{\mathcal R} = \int_{-\infty}^\infty \exp(-r\tau) f_{\tiny\textrm{exp}}(\tau) \mathrm{d} \tau,
\end{equation}
where the forward serial-interval distribution during the exponential growth phase is defined as:
\begin{equation}
f_{\tiny\textrm{exp}}(\tau) \propto \int_{0}^\infty \int_{0}^\infty \exp( -r x_0) h(x_0, \sigma) k(\tau-\sigma+x_0) \mathrm{d} x_0\, \mathrm{d}\sigma.
\end{equation}
In Appendix, we provide a mathematical proof that this relationship holds.

@@

We use a simulation-based approach to compare the estimates of $\mathcal R$ based on the serial- and generation-interval distributions. 
To do so, we model the intrinsic generation-interval distribution and the incubation period using a multivariate log-normal distribution with log means $\mu_G, \mu_I$, log standard variances $\sigma_G^2, \sigma_I^2$, and correlation $\rho$.
The multivariate log-normal distribution is parameterized based on pameter estimates for COVID-19.
We construct the forward serial-interval distribution during the exponential growth period as follows:
\begin{equation}
S_i = -B_i + (G_i|B_i) + I_i,
\end{equation}
where the backward incubation period $B_i$ of an infector is simulated by drawing random log-normal samples $A_i$ with log mean $\mu_I$ and log variance $\sigma_I^2$ and resampling $A_i$, each weighted by the inverse of the exponential growth function $\exp(-rA_i)$;
the intrinsic generation interval conditional on the incubation period of the infector $(G_i|B_i)$ is drawn from a log-normal distribution with log mean $\mu_G + \sigma_G \rho (\log(B_i) - \mu_I)/\sigma_I$ and log variance $\sigma_G^2 (1-\rho^2)$;
the forward incubation period $I_i$ of an infectee is drawn from a log-normal distribution with log mean $\mu_I$ and log variance $\sigma_I^2$.
We then calculate the reproduction number $\mathcal R$ using the empirical estimator:
\begin{equation}
\mathcal R = \frac{1}{\frac{1}{N}\sum_{i=1}^N \exp(- r S_i)}.
\end{equation}
We compare this estimate with a naive estimate of $\mathcal R$ that assumes that the backward and the forward incubation periods are identically distributed:
\begin{equation}
\mathcal R = \frac{1}{\frac{1}{N}\sum_{i=1}^N \exp(- r T_i)},
\end{equation}
where
\begin{equation}
T_i = -A_i + (G_i|A_i) + I_i.
\end{equation}

@@

Recall that the forward serial interval can be written as:
\begin{equation}
- x_0 + \sigma + x_1
\end{equation}
Note that $x_1$ is independent of $x_0$ and $\sigma$. Then, we get:
\begin{equation}
M_{- x_0 + \sigma + x_1}(-r) = M_{- x_0 + \sigma}(-r) M_{x_1}(-r).
\end{equation}
We want to show that 
\begin{equation}
M_{- x_0 + \sigma}(-r)= \frac{M_\sigma(-r)}{M_{x_1}(-r)}
\end{equation}
for $r \geq 0$.
Note that the time between symptom onset and infection of an infectee follows the following distribution:
\begin{equation}
z(\tau) \propto \int_{\max(0, -\tau)}^\infty i(\ell - x_0) h_{\ell - x_0}(x_0, \tau+x_0) \mathrm{d} x_0.
\end{equation}
During the exponential growth phase, we have
\begin{equation}
\begin{aligned}
z_{\tiny\textrm{exp}}(\tau) &= \frac{1}{N} \int_{\max(0, -\tau)}^\infty \exp(- r x_0) h(x_0, \tau+x_0) \mathrm{d} x_0\\
% &= \frac{1}{N} \int_{0}^\infty \left[ \exp(- r x_0) k(x_0) \times \frac{h(x_0, \tau+x_0)}{k(x_0)}\right] \mathrm{d} x_0
\end{aligned}
\end{equation}
where $N$ is the normalization factor:
\begin{equation}
\begin{aligned}
N &= \int_{-\infty}^\infty \int_{\max(0, -\tau)}^\infty \exp(- r x_0) h(x_0, \tau+x_0) \mathrm{d} x_0\,\mathrm{d}\tau\\
&= \int_{0}^\infty \int_{-x_0}^\infty \exp(- r x_0) h(x_0, \tau+x_0) \mathrm{d}\tau\,\mathrm{d} x_0\\
&= \int_{0}^\infty \exp(- r x_0) k(x_0) dx_0\\
&= M_{x_1}(-r)
\end{aligned}
\end{equation}
Therefore,
\begin{equation}
\begin{aligned}
&\int_{-\infty}^{\infty} \exp(-r\tau) z_{\tiny\textrm{exp}}(\tau) \mathrm{d}\tau\\
&=\frac{1}{M_{-x_1}(-r)} \int_{-\infty}^{\infty} \exp(-r\tau) \int_{\max(0, -\tau)}^\infty \exp(- r x_0) h(x_0, \tau+x_0) \mathrm{d} x_0\, \mathrm{d}\tau
\end{aligned}
\end{equation}

>>>>

Since the emergence of the novel coronavirus disease (COVID-19), a significant amount of research has focused on estimating its basic reproduction number $\mathcal R_0$ \citep{majumder2020early}.
The basic reproduction number, which is defined as the average number of secondary cases caused by a primary case in a fully susceptible population, allows us to predict the extent to which a disease will spread in the population and the amount of intervention to prevent an outbreak \citep{anderson1991infectious}.
Since the reproduction number cannot be measured directly, it is often estimated from the observed exponential growth rate using generation- and serial-interval distributions.

The generation interval is defined as the time between when an individual (infector) is infected and when an individual infects another person (infectee);
the generation-interval distributions plays a key role in shaping the relationship between the exponential growth rate $r$ and the reproduction number $\mathcal R$.
On the other hand, the serial interval is defined as the time between when an infector and an infectee become symptomatic.
Previous studies have often alluded to differences in their variances, despite having seemingly identical expected values, and noted that using serial-interval distribution can give different estimates of $\mathcal R$.
Even though these distinctions were first made over a decade ago \citep{svensson2007note}, 
the need for a better conceptual and theoretical framework for understanding their differences is becoming clearer as the COVID-19 pandemic unfolds:
Researchers continue to rely on both generation and serial intervals to estimate $\mathcal R$ for COVID-19 without making a clear distinction, and some even mislabel them.

@@

During the initial phase of an epidemic, the proprotion susceptible remains constant ($S(t) = S(0)$) and incidence of infection grows exponentially: $i(t)=i_0\exp(rt)$.
Then, we can estimate the reproduction number from the exponential growth rate $r$ via the Euler-Lotka equation:
\begin{equation}
\frac{1}{\mathcal R} = \int_0^\infty \exp(-r\tau) g(\tau) \mathrm{d} \tau.
\end{equation}
Like forward generation-interval distributions, 
forward serial-interval distributions describe the renewal process of symptomatic cases.
Therefore, the forward serial-interval distribution $f_{\textrm{\tiny exp}}(\tau)$ during the exponential growth phase provides the identical $r$--$\mathcal R$ link as the intrinsic generation-interval distribution:
\begin{equation}
\frac{1}{\mathcal R} = \int_{-\infty}^\infty \exp(-r\tau) f_{\textrm{\tiny exp}}(\tau) \mathrm{d} \tau,
\end{equation}
where the forward serial-interval distribution during the exponential growth phase is defined as:
\begin{equation}
f_{\textrm{\tiny exp}}(\tau) \propto \int_{0}^\infty \int_{0}^\infty \exp( -r x_0) h(x_0, \sigma) k(\tau-\sigma+x_0) \mathrm{d} x_0\, \mathrm{d}\sigma.
\end{equation}
In Appendix, we provide a mathematical proof that this relationship holds.

@@

We use a simulation-based approach to compare the estimates of $\mathcal R$ based on the serial- and generation-interval distributions. 
To do so, we model the intrinsic generation-interval distribution and the incubation period using a multivariate log-normal distribution with log means $\mu_G, \mu_I$, log standard variances $\sigma_G^2, \sigma_I^2$, and correlation $\rho$.
The multivariate log-normal distribution is parameterized based on pameter estimates for COVID-19.
We construct the forward serial-interval distribution during the exponential growth period as follows:
\begin{equation}
S_i = -B_i + (G_i|B_i) + I_i,
\end{equation}
where the backward incubation period $B_i$ of an infector is simulated by drawing random log-normal samples $A_i$ with log mean $\mu_I$ and log variance $\sigma_I^2$ and resampling $A_i$, each weighted by the inverse of the exponential growth function $\exp(-rA_i)$;
the intrinsic generation interval conditional on the incubation period of the infector $(G_i|B_i)$ is drawn from a log-normal distribution with log mean $\mu_G + \sigma_G \rho (\log(B_i) - \mu_I)/\sigma_I$ and log variance $\sigma_G^2 (1-\rho^2)$;
the forward incubation period $I_i$ of an infectee is drawn from a log-normal distribution with log mean $\mu_I$ and log variance $\sigma_I^2$.
We then calculate the reproduction number $\mathcal R$ using the empirical estimator:
\begin{equation}
\mathcal R = \frac{1}{\frac{1}{N}\sum_{i=1}^N \exp(- r S_i)}.
\end{equation}
We compare this estimate with a naive estimate of $\mathcal R$ that assumes that the backward and the forward incubation periods are identically distributed:
\begin{equation}
{\mathcal R}_{\textrm{\tiny naive}} = \frac{1}{\frac{1}{N}\sum_{i=1}^N \exp(- r T_i)},
\end{equation}
where
\begin{equation}
T_i = -A_i + (G_i|A_i) + I_i.
\end{equation}

@@

Recall that the forward serial interval can be written as:
\begin{equation}
- x_0 + \sigma + x_1
\end{equation}
Note that $x_1$ is independent of $x_0$ and $\sigma$. Then, we get:
\begin{equation}
M_{- x_0 + \sigma + x_1}(-r) = M_{- x_0 + \sigma}(-r) M_{x_1}(-r).
\end{equation}
We want to show that 
\begin{equation}
M_{- x_0 + \sigma}(-r)= \frac{M_\sigma(-r)}{M_{x_1}(-r)}
\end{equation}
for $r \geq 0$.
Note that the time between symptom onset and infection of an infectee follows the following distribution:
\begin{equation}
z(\tau) \propto \int_{\max(0, -\tau)}^\infty i(\ell - x_0) h_{\ell - x_0}(x_0, \tau+x_0) \mathrm{d} x_0.
\end{equation}
During the exponential growth phase, we have
\begin{equation}
\begin{aligned}
z_{\textrm{\tiny exp}}(\tau) &= \frac{1}{N} \int_{\max(0, -\tau)}^\infty \exp(- r x_0) h(x_0, \tau+x_0) \mathrm{d} x_0\\
% &= \frac{1}{N} \int_{0}^\infty \left[ \exp(- r x_0) k(x_0) \times \frac{h(x_0, \tau+x_0)}{k(x_0)}\right] \mathrm{d} x_0
\end{aligned}
\end{equation}
where $N$ is the normalization factor:
\begin{equation}
\begin{aligned}
N &= \int_{-\infty}^\infty \int_{\max(0, -\tau)}^\infty \exp(- r x_0) h(x_0, \tau+x_0) \mathrm{d} x_0\,\mathrm{d}\tau\\
&= \int_{0}^\infty \int_{-x_0}^\infty \exp(- r x_0) h(x_0, \tau+x_0) \mathrm{d}\tau\,\mathrm{d} x_0\\
&= \int_{0}^\infty \exp(- r x_0) k(x_0) dx_0\\
&= M_{x_1}(-r)
\end{aligned}
\end{equation}
Therefore,
\begin{equation}
\begin{aligned}
&\int_{-\infty}^{\infty} \exp(-r\tau) z_{\textrm{\tiny exp}}(\tau) \mathrm{d}\tau\\
&=\frac{1}{M_{-x_1}(-r)} \int_{-\infty}^{\infty} \exp(-r\tau) \int_{\max(0, -\tau)}^\infty \exp(- r x_0) h(x_0, \tau+x_0) \mathrm{d} x_0\, \mathrm{d}\tau
\end{aligned}
\end{equation}
